Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-02-04 _paper.rmd
-regression use case: fix rotation of basis. -screen captures relabelled and text adjusted.

<!-- Local explanations as par coords -->
The local explanations for all observations are normalized (squared sum of values adds to 1) and thus the relative importance of features can be compared across all instances. These are depicted as a vertical parallel coordinate plot, where each line connects one instance's feature attribution (Fig. \@ref(fig:classificationcase) and \@ref(fig:regressioncase) d). The attribution projections of the PI and CI are shown as dashed and dotted lines, respectively. From this plot, we would read the range of importances across all instances can be interpreted. For classification, one would look at differences between groups on any feature, for example, Fig. \@ref(fig:classificationcase) d suggests that `b_l` is important for distinguishing the green class from the other two. For regression, one might generally observe which features have low values for all instances (not important), for example, `BMI` and `pwr` in \@ref(fig:regressioncase) d, and which have a range of high and low values (e.g. `off`, `def`) suggesting important for some instances and not important for other instances.

@@

<!-- Tour animation -->
The attribution projection of the PI is the initial 1D basis in a radial tour, displayed as a density plot for a categorical response (Fig. \@ref(fig:classificationcase), e), and as scatterplots for a quantitative response ((Fig. \@ref(fig:regressioncase), e). The PI and CI are indicated by vertical dashed and dotted lines, respectively. The user uses the radial tour to vary contribution of the selected feature between 0-1. Doing so tests the sensitivity of structure (class separation or strength of relationship) to the feature's contribution. For classification, if the separation between classes diminishes when the feature contribution is reduced then this suggests that the feature is important for the class separation. For regression, if the relationship scatterplot weakens when the feature contribution is reduced then this suggests that the feature is important for accurately predicting the response.

<!-- The default feature selected has the largest discrepancy between the attribution of the PI and CI. The data (scaled by standard deviation away from the mean) is projected through the current basis --- the horizontal position of Fig. \@ref(fig:classificationcase) e and Fig. \@ref(fig:regressioncase) d.-->

@@

Selecting a misclassified instance as PI and a correctly classified point nearby in data space as CI, makes it easier to examine the features most responsible for the error. The global view (right plot) displays the model confusion matrix. The radial tour is 1D, plotted as a density where colour indicates class. A scrollbar here enables the user to vary the contribution of each feature to explore the sensitivity the separation to that feature.

```{r classificationcase, echo=F, out.width = "100%", fig.align="center", fig.cap = "Overview of the cheem viewer for classification tasks. Global view inputs, (a), set the PI, CI, and color statistic. Global view, (b) PC1 by PC2 for data space, attribution space, and prediction by observed y (visual of the confusion matrix for classification). Points are colored by predicted class, and red circles indicate misclassified instances. Radial tour inputs (c) select features to include and which feature is changed in the tour. The visual depicting 1D basis (d) shows the distribution of the feature attribution, and bars show the current basis. The black bar is being varied. Panel (e) is the resulting projection of the data indicated as density in the classification case."}
knitr::include_graphics("./figures/app_classification.PNG")
```

@@

Selecting an inaccurately predicted instance as PI and an accurately predicted instance, with similar feature values, as CI is a useful way to understand how the model is failing, or not. The global view (right plot) shows a scatterplot of the observed vs predicted values, which should exhibit a strong relationship if the model is a good fit. The points can be colored by a statistic, residual, a measure of outlyingness (log Mahalanobis distance) or correlation, to help with understand where the model fits better or worse.

@@

```{r regressioncase, echo=F, out.width="100%", fig.cap="Overview of the cheem viewer for regression task highlighting the differences from the classification task and interactive features. Panel (a) shows linked brushing across the global view and the tooltip display when the cursor hovers over an instance. Coloring on a statistic (b) highlights structure organized in the attribution space. Interactive tabular display (d) populates when instances are selected. Regression projection (e) uses the same horizontal projection and fixes the vertical positions to the observed y and residuals, (left and right)."}
knitr::include_graphics("./figures/app_regression_interactions.PNG")
```

@@

With figure \@ref(fig:casefifa), we will test the premise of the local explanation. While offensive and reaction skills (`off` and `rct`) are both crucial to explaining a star offensive player. As the contribution of defensive skills increases, Messi's is no longer separated from the group, and other defensive players are better predicted in this attribution case. In terms of what-if analysis, his predicted wages would be halved if Messi's tree SHAP attributions at these levels.

>>>>

<!-- Local explanations as par coords -->
The local explanations for all observations are normalized (squared sum of values adds to 1) and thus the relative importance of features can be compared across all instances. These are depicted as a vertical parallel coordinate plot, where each line connects one instance's feature attribution (Fig. \@ref(fig:classificationcase) e and \@ref(fig:regressioncase) e). The attribution projections of the PI and CI are shown as dashed and dotted lines, respectively. From this plot, we would read the range of importances across all instances can be interpreted. For classification, one would look at differences between groups on any feature, for example, Fig. \@ref(fig:classificationcase) e suggests that `b_l` is important for distinguishing the green class from the other two. For regression, one might generally observe which features have low values for all instances (not important), for example, `BMI` and `pwr` in \@ref(fig:regressioncase) e, and which have a range of high and low values (e.g. `off`, `def`) suggesting important for some instances and not important for other instances.

@@

<!-- Tour animation -->
The attribution projection of the PI is the initial 1D basis in a radial tour, displayed as a density plot for a categorical response (Fig. \@ref(fig:classificationcase), f), and as scatterplots for a quantitative response (Fig. \@ref(fig:regressioncase), f). The PI and CI are indicated by vertical dashed and dotted lines, respectively. The user uses the radial tour to vary contribution of the selected feature between 0-1. Doing so tests the sensitivity of structure (class separation or strength of relationship) to the feature's contribution. For classification, if the separation between classes diminishes when the feature contribution is reduced then this suggests that the feature is important for the class separation. For regression, if the relationship scatterplot weakens when the feature contribution is reduced then this suggests that the feature is important for accurately predicting the response. <!-- The default feature selected has the largest discrepancy between the attribution of the PI and CI. The data (scaled by standard deviation away from the mean) is projected through the current basis --- the horizontal position of Fig. \@ref(fig:classificationcase) f and Fig. \@ref(fig:regressioncase) f.-->

@@

Selecting a misclassified instance as PI and a correctly classified point nearby in data space as CI, makes it easier to examine the features most responsible for the error. The global view (Fig. \@ref(fig:classificationcase) c) displays the model confusion matrix. The radial tour is 1D, plotted as a density where color indicates class. A scroll bar here enables the user to vary the contribution of each feature to explore the sensitivity the separation to that feature.

```{r classificationcase, echo=F, out.width = "100%", fig.align="center", fig.cap = "Overview of the cheem viewer for classification tasks. Global view inputs, (a), set the PI, CI, and color statistic. Global view, (b) PC1 by PC2 approximations of the data space and attribution space. (c) prediction by observed y (visual of the confusion matrix for classification tasks). Points are colored by predicted class, and red circles indicate misclassified instances. Radial tour inputs (d) select features to include and which feature is changed in the tour. (e) shows parallel coordinate display of the distribution of the feature attributions while bars dipict contribution for the current basis. The black bar is the variable being changed in the radial tour. Panel (f) is the resulting projection of the data indicated as density in the classification case."}
knitr::include_graphics("./figures/app_classification.PNG")
```

@@

Selecting an inaccurately predicted instance as PI and an accurately predicted instance, with similar feature values, as CI is a useful way to understand how the model is failing, or not. The global view (Fig. \@ref(fig:regressioncase) a) shows a scatterplot of the observed vs predicted values, which should exhibit a strong relationship if the model is a good fit. The points can be colored by a statistic, residual, a measure of outlyingness (log Mahalanobis distance) or correlation, to help with understand where the model fits better or worse.

@@

```{r regressioncase, echo=F, out.width="100%", fig.cap="Overview of the cheem viewer for regression task highlighting the differences from the classification task and interactive features. Panel (a) shows linked brushing across the global view and the tooltip display when the cursor hovers over an instance. Applied linked brushing, (b), selected points are highlighted in the component spaces and tabularly displayed. Coloring on a statistic (c) highlights structure organized in the attribution space. Interactive tabular display (d) populates when instances are selected. Contribution of the 1D basis effecting the horizontal position (e) parallel coorindate display of the feature attribution from all observations, and horizontal bars show the contribution to the current basis. Regression projection (f) uses the same horizontal projection and fixes the vertical positions to the observed y and residuals, (left and right)."}
knitr::include_graphics("./figures/app_regression_interactions.PNG")
```

@@

With figure \@ref(fig:casefifa), we will test the premise of the local explanation. While offensive and reaction skills (`off` and `rct`) are both crucial to explaining a star offensive player. As the contribution of defensive skills increases, Messi's is no longer separated from the group, and the defensive players are better predicted in this attribution case. Essentially, the further right an observation is the better the the basis does at describing its predicted value. In terms of what-if analysis, his predicted wages would be halved if Messi's tree SHAP attributions at these levels.
